Widespread nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2023-06-27 manuscript/manuscript_file.Rmd
Reworked conclusion

While NO~3~ and water temperature were the primary drivers of N~2~O production or consumption rates, waterbody morphology determined the extent to which N~2~O processing affected dissolved N~2~O in surface waters. When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreases with increasing waterbody size class **[Fig. 4A]**. This pattern holds regardless of whether N~2~O is supersaturated or undersatured **[Fig. 4B, C]**.  Waterbody size and shape can affect dissolved gas concentration through several mechanisms. In lakes with low water volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column (@carpenter_1983) . Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep lakes, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface (@Beaulieu_etal_2015, @Beaulieu_etal_2014).  Unfortunately, the morphometric data required to test these hypothesis (e.g. mean depth, max depth) are not available for the samples waterbodies.

Persistent patterns in the air-water gas exchange rate across lake size classes may also contribute to greater N~2~O disequlibrim in smaller waterbodies. Air-water gas exchange will vent excess N~2~O from surface waters or facilitate the invasion of air that can replenish depleted N~2~O in undersaturated waterbodies. Regardless of whether a waterbody is superaturated or undersatured in N~2~O, air-water gas exchange will function to move the dissolved N~2~O concentration toward equilibrium. One of the primary drivers of air-water gas exchange in lentic waters is wind speed (@Vachon_etal_2013). Wind speeds tend to be greater in large lakes with long fetches and lower in small waterbodies, particularly if they are sheltered by shoreline vegetation. Thus, greater air-water gas exchange rates in large lakes may maintain dissolved N~2~O concentrations close to equilibrium, whereas the affect of biogeochemistry on dissolved N~2~O is more pronounced in small waterbodies because low air-water gas exchange limits the rate at which excess N~2~O can be ventilated or depleted N~2~O can be replenished.

@@

## Fluxes 
Total N~2~O flux from the surface of CONUS waterbodies is a function of the N~2~O emission rate and waterbody surface area.  The largest lake size class (> 50 Ha) had the greatest N~2~O flux in all ecoregions, but flux estimates for this lake size class were also the most uncertain (Fig. 6). Small uncertainties in emission rates translate to large uncertainties in flux when expressed across the surface area of large waterbodies.  Despite widespread N~2~O undersaturation, all ecoregions were a net source of N~2~O to the atmosphere (Table 1).  Differences in N~2~O flux across ecoregions are explained by patterns in N~2~O saturation and waterbody surface.  The Coastal Plains ecoregion had the greatest flux, driven by the high levels of N~2~O saturation and extensive waterbody surface area. By contrast, the Northern Plains had the lowest N~2~O saturation, waterbody surface area, and N~2~O flux.

At the national scale, we estimate that CONUS waterbodies emit [problem knitting inline code? table1_dat %>% filter(`spatial domain` == "national") %>% select(`N2O flux`) %>% pull()] 1049 (358 - 1581) Mg N~2~O year^{-1}, assuming that the emission rates reported here hold for an entire calendar year. This is likely a conservative estimate because our data is based on summer conditions, whereas several studies have reported maximum N~2~O emission rates during the winter (citation, see climate benefits paper) and Fall (harsha n2o paper, LD n2o paper) in lakes and reservoirs.  While we are unaware of independent estimates of N~2~O emissions from CONUS lakes and reservoirs, the Intergovernmental Panel on Climate Change estimates that CONUS drainage networks, inclusive of zero order streams through major river mouths, receiving run off from croplands and grasslands emit `r ipcc.n2o.surface` (22,767) Mg N~2~O year^{-1} (@epa_2023).  Emissions from lakes and reservoirs represent only `r round((national_stats %>% select(post_med_fn2o) %>% pull() / ipcc.n2o.surface) * 100, digits = 1)` % of total surface water emissions, with the balance emanating from streams and rivers.  Emissions from CONUS lakes and reservoirs are also small when compared to other anthropogenic N~2~O sources including wastewater treatment (88,00  Mg N~2~O year^{-1}), stationary combustion (84,000 Mg N~2~O year^{-1}), and manure management (66,000  Mg N~2~O year^{-1}).

@@

## Conclusion
* nearly three quarters of CONUS lakes and reservoirs are N2O sinks during the summer!
* despite widespread undersaturation, CONUS lakes and waterbodies are a small source of N2O
* our study is limited in temporal scope, would seasonality change the story?
* greatest uncertainty in flux is for large lakes.  Focused investigations on the nation's largest waterbodies to improve estimate?

>>>>

While NO~3~ and water temperature were the primary drivers of N~2~O production or consumption rates, dissolved N~2~O concentration was also related to waterbody size.  When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreases with increasing waterbody size class **[Fig. 4A]**. This pattern holds regardless of whether N~2~O is supersaturated or undersatured **[Fig. 4B, C]**.  Waterbody size and shape can affect dissolved gas concentration through several mechanisms. In lakes with low water volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column (@carpenter_1983). Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep lakes, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface (@Beaulieu_etal_2015, @Beaulieu_etal_2014).

Persistent patterns in the air-water gas exchange rate across lake size classes may also contribute to greater N~2~O disequilibrium in smaller waterbodies. Air-water gas exchange will vent excess N~2~O from surface waters or facilitate the invasion of air that can replenish depleted N~2~O in undersaturated waterbodies. Regardless of whether a waterbody is superaturated or undersatured in N~2~O, air-water gas exchange will function to move the dissolved N~2~O concentration toward equilibrium. One of the primary drivers of air-water gas exchange in lentic waters is wind speed (@Vachon_etal_2013). Wind speeds tend to be greater in large lakes with long fetches and lower in small waterbodies, particularly if they are sheltered by shoreline vegetation. Thus, greater air-water gas exchange rates in large lakes may maintain dissolved N~2~O concentrations close to equilibrium, whereas the affect of biogeochemistry on dissolved N~2~O is more pronounced in small waterbodies because low air-water gas exchange limits the rate at which excess N~2~O can be ventilated or depleted N~2~O can be replenished.

@@

## Fluxes 
Total N~2~O flux from the surface of CONUS waterbodies is a function of the N~2~O emission rate and waterbody surface area.  The largest lake size class (> 50 Ha) had the greatest N~2~O flux in all ecoregions, but flux estimates for this lake size class were also the most uncertain (Fig. 6). Small uncertainties in emission rates translate to large uncertainties in flux when expressed across the surface area of large waterbodies.  Despite widespread N~2~O undersaturation, all ecoregions were a net source of N~2~O to the atmosphere (Table 1).  Differences in N~2~O flux across ecoregions are explained by patterns in N~2~O saturation and waterbody surface (Table 1).  The Coastal Plains ecoregion had the greatest flux, driven by the high levels of N~2~O saturation and extensive waterbody surface area. By contrast, the Northern Plains had the lowest N~2~O saturation, waterbody surface area, and N~2~O flux.

At the national scale, we estimate that CONUS waterbodies emit [problem knitting inline code? table1_dat %>% filter(`spatial domain` == "national") %>% select(`N2O flux`) %>% pull()] 1049 (358 - 1581) Mg N~2~O year^{-1}, assuming that the emission rates reported here hold for an entire calendar year. This is likely a conservative estimate because our data is based on summer conditions, whereas several studies have reported maximum N~2~O emission rates during the winter (citation, see climate benefits paper) and Fall (harsha n2o paper, LD n2o paper) in lakes and reservoirs.  While we are unaware of independent estimates of N~2~O emissions from CONUS lakes and reservoirs, the Intergovernmental Panel on Climate Change estimates that CONUS drainage networks, inclusive of zero order streams through major river mouths, receiving run off from croplands and grasslands emit `r ipcc.n2o.surface` (22,767) Mg N~2~O year^{-1} (@epa_2023).  Emissions from lakes and reservoirs represent only `r round((national_stats %>% select(post_med_fn2o) %>% pull() / ipcc.n2o.surface) * 100, digits = 1)` % of total surface water emissions reported by the IPCC (@epa_2023), with the balance emanating from streams and rivers.  Emissions from CONUS lakes and reservoirs are also small when compared to other anthropogenic N~2~O sources including wastewater treatment (88,00  Mg N~2~O year^{-1}), stationary combustion (84,000 Mg N~2~O year^{-1}), and manure management (66,000  Mg N~2~O year^{-1}).

@@

## Conclusion
Our survey is the first to demonstrate that most lakes and reservoirs in the CONUS function as N~2~O sinks during the summer.  Despite widespread N~2~O uptake, CONUS waterbodies function as a net source of N~2~O due to high emissions from from a subset of the population.  Dissolved N~2~O in lakes and reservoirs can be predicted from non-linear interactions between water temperature, NO~3~ availability, and waterbody size.  NO~3~ availability largely determines whether a waterbody functions as a sink or source of N~2~O, and the magnitude of the source or sink is exaggerated with increasing water temperature. The degree of N~2~O disequilibrium also increases with decreasing waterbody size, possibly because smaller waterbodies have low air-water gas exchange rates.

Although waterbodies in the largest lake size class tend to have N~2~O levels close to equilibrium, they emit more N~2~O than the other size classes due to their expansive surface area.  The N~2~O flux from these lakes is also more uncertain than the smaller lake size classes because relatively small uncertainties in dissolved N~2~O translate to large uncertainties in flux when propagated across the surface of large waterbodies.  Efforts to improve the accuracy of the lentic N~2~O flux estimate for the U.S. should therefore focus on investigations of the nations largest lakes and reservoirs.

The smallest lake size classes exhibited both the highest and lowest levels of N~2~O saturation, depending on NO~3~ availability.  It is unclear whether this variability is related to differences in NO~3~ loading or temporal variation.  For example, reservoirs have been shown to function as N~2~O sources in early summer when NO~3~ is replete, but switch to N~2~O sinks in later summer when the NO~3~ supply has been exhausted (@beaulieu_etal_2014).  A better understanding of the temporal evolution of N~2~O production and consumption in small waterbodies is an important research need to determine the extent to which these systems can remove excess nitrogen from surface waters and simulatenously mitigate N~2~O emissions.
